---
title: "imprintome_vs_wgbs_methylation"
output: 
  html_document:
    toc: true # table of content true
    number_sections: true  ## if you want number sections at each table header
    highlight: tango  # specifies the syntax highlighting style
    theme: united 
    toc_float: true
    code_folding: hide
    toc_depth: 3  # upto three depths of headings (specified by #, ## and ###)
vignette: >
  %\VignetteIndexEntry{imprintome_vs_wgbs_methylation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
devtools::load_all()
# detach(name = "package:tdhia", unload=TRUE)
# remove.packages("tdhia")
# eblvm_path <- devtools::build(pkg = paste0(dirname(here::here()), "/eblvm"))
# install.packages(eblvm_path, lib='eblvm',repos = NULL, type="source",clean = TRUE)

```



This vignette will demonstrate how to load and process the imprintome files from the True Diagnostic Human Imprintome Array, along with conducting several statistical analysis on the data. We will use a dataset from:


> Carreras-Gallo N et al. Creation and Validation of the First Infinium DNA Methylation Array for the Human Imprintome. bioRxiv [Preprint]. 2024 Jan 16:2024.01.15.575646. doi: 10.1101/2024.01.15.575646. PMID: 38293193; PMCID: PMC10827131.


Specifically, we will replicate and extend the results from Figure 4. The dataset is comprised of 16 patients, half of which have varying degrees of Alzheimer's disease (N=8), and the other half are controls (N=8). Each patient has their imprintome methylation status measured from whole blood with whole genome bisulfite sequencing (WGBS) and from whole blood and the imprintome array.


# Load packages and helper functions

We start by loading required packages

1. tdhia: True Diagnostic Human Imprintome Array package.

2. rappdirs: data storage path management.

3. tidyverse: data processing and plotting. 

4. cowplot: plotting and saving plots to disk.


```{r setup}

library(tdhia)
library(rappdirs)
library(tidyverse)
library(cowplot)
```


We also load two user-defined helper functions for orgnaizing data to compare imprintome to wgbs data.

1. *group_by_patient()*: For each patient, group beta/ methylation % values from a specific imprintome pipeline, along with methylation from WGBS.

2. *group_by_mean()*: For study group, group mean beta/ methylation % values from a specific imprintome pipeline, along with methylation from WGBS.


```{r setup2}

#' group_by_patient
#' @description For each patient, group beta/ methylation % values from a 
#' specific imprintome pipeline, along with methylation from WGBS.
#' 
#' @param df_study dataframe of study data, contains data for each patient to 
#' imprintome vs wgbs.
#' @param df_cpg_imp methylation data from a particular imprintome pipeline
#' @param df_cpg_wgbs methylation data from wgbs data
#' @param col_label string column label, used to label name of pipeline
#' 
#' @return
group_by_patient <- function(df_study, df_cpg_imp, df_cpg_wgbs, col_label) {
  
  df_list <- list()
  for (n in 1:nrow(df_study)) {
    # colnames(df_cpg_imp == df_study$Patient.ID[n]
    x = select(df_cpg_imp, one_of(df_study$Patient.ID[n])) %>% rownames_to_column("cpg_id")
    y = select(df_cpg_wgbs, one_of(df_study$wgbs_base_names[n])) %>% rownames_to_column("cpg_id")
    df_join <- left_join(x = x, y = y, by = "cpg_id", keep = FALSE, na_matches = "never",
                         unmatched= "drop", relationship = "one-to-one")
    colnames(df_join) <- c("cpg_id", "imp","wgbs")
    df_join$sample_id <- df_study$wgbs_base_names[n]
    df_join$name <- col_label
    df_list[[n]] <- df_join
  }
  df_comp  <- do.call(rbind, df_list)
  return(df_comp)
}


#' group_by_mean
#' @description For study group, group mean beta/ methylation % values from a 
#' specific imprintome pipeline, along with methylation from WGBS.
#' 
#' @param df_study dataframe of study data, contains data for each patient to 
#' imprintome vs wgbs.
#' @param df_cpg_imp methylation data from a particular imprintome pipeline
#' @param df_cpg_wgbs methylation data from wgbs data
#' @param col_label string column label, used to label name of pipeline
#' 
#' @return
group_by_mean <- function(df_study, df_cpg_imp, df_cpg_wgbs, col_label, patient_id = 0) {
  
  # Get unique study groups to average patients from each groups
  unq_study_group <- unique(df_study$study_group)
  
  df_list <- list()
  for (n in 1:length(unq_study_group)) {
    # Binary index of patients that match group
    bind <- df_study$study_group == unq_study_group[n]
    
    
    x <- data.frame(imp = rowMeans(select(df_cpg_imp, df_study$Patient.ID[bind]))) %>%
      rownames_to_column("cpg_id")
    y <- data.frame(wgbs = rowMeans(select(df_cpg_wgbs, df_study$wgbs_base_names[bind]))) %>% 
      rownames_to_column("cpg_id")
    
    df_join <- left_join(x = x, y = y, by = "cpg_id", keep = FALSE, na_matches = "never",
                         unmatched= "drop", relationship = "one-to-one")
    df_join$sample_id <- unq_study_group[n]
    df_join$pipeline_name <- col_label
    df_join$patient_id <- patient_id 
    df_list[[n]] <- df_join
  }
  df_comp  <- do.call(rbind, df_list)
  return(df_comp)
}

mft <- tdhia::manifest_v1A2_design_scores
```

# Load and clean study metadata

Study metadata are read from several csv files that join together info several pieces of information.

1. WGBS filnames to study_ids.

2. Imprintome filenames to study_ids.

3. Study_ids to demographical information about the study participant.

All of this information is collected in the study_metadata dataframe.

```{r metadata, include = TRUE, cache = FALSE}
# Define local path variables
idat_dir_path <- system.file("extdata", package = "tdhia")
out_path = paste0(dirname(getwd()), "/data/tdhia_brain")
# Loading AD study data 
#_______________________________________________________________________________
pdata_wgbs <- tdhia::alzheimer_study_wbgs_shared_cpg_metadata
pdata_imprintome <- read.csv(file = paste0(idat_dir_path, "/Imprintome_NC_State_pData.csv"))
# Filtering metadata rows
pdata_imprintome <- pdata_imprintome[grepl(pattern = "[A-Z]{2}", x = pdata_imprintome$Sample),]
pdata_imprintome$sampleID <- as.numeric(stringr::str_extract(pdata_imprintome$Sample, "[0-9]+$"))
pdata_wgbs_full <- 
  dplyr::left_join(x = pdata_wgbs, y = pdata_imprintome, by = "sampleID", keep = FALSE) %>% 
  dplyr::filter(!is.na(Patient.ID))
# Get basenames from wgbs file names
pdata_wgbs_full$wgbs_base_names <- sapply(strsplit(pdata_wgbs_full$FileName, "_"),
                                          function(x) x[1])

# Add patient id as rownames
rownames(pdata_wgbs_full) <- pdata_wgbs_full$Patient.ID

wgbs_complete <- read.csv(file = paste0(system.file("data-raw", package = "tdhia"), "/AD.WGBS.Brain.samples.TimeFrozen.csv"))

# merge with additional study data
study_metadata <- left_join(x = pdata_wgbs_full, y = wgbs_complete %>% dplyr::rename(
  sampleID = Sample.ID), by = "sampleID", keep = FALSE, unmatched = "drop")

study_metadata$Treatment <- factor(study_metadata$Treatment)
study_metadata$AD.Case.Control <- factor(study_metadata$AD.Case.Control)
study_metadata$Ethnicity <- factor(study_metadata$Ethnicity)
study_metadata$Age <- as.numeric(str_replace(string = study_metadata$Age, pattern = ">", ""))
study_metadata$Sex <- factor(study_metadata$Sex)
rownames(study_metadata) <- study_metadata$Patient.ID


# COnvert variables needed for statisticla analysis to factors or boolean variables
study_metadata$race <-  study_metadata$Ethnicity
study_metadata$is_control = study_metadata$AD.Case.Control == "Control"
study_metadata$disease_state = study_metadata$AD.Case.Control != "Control"
study_metadata$study_group <- study_metadata$Treatment


```


# Process Imprintome with Various Settings and Pair with WGBS data

1. **default**: beta values from standard default pipeline. The SeSame R package is used for processing the methylation data. Raw IDAT files are loaded, the Infinuium probe type is inferred, flourescent signal is subtracted with Norm-Exp deconvolution that is parameterized with out-of-band probes, and a non-linear dye bias correction is applied.

2. **quant_norm**: beta values from standard pipeline, but performed quantile normalization (patient-wise) during cpg conversion step.

3. **cpg_smooth**: beta values from standard pipeline, but cpg smoothing is performed (moving average of beta values from adjacent sites within the same ICR).
oocyte

4. **low_p_val**: beta values from standard pipeline, but using a strict p-value threshold of 0.05 for sesame signal detection p-value instead of the usual.

5. **design_thresh**: exclude probes that have an illumina probe design score below 0.5.

6. **pre_merge**: beta values from standard pipeline, but replicate probes are averaged *before beta values are calculated*, meaning the averaged by methylation and unmethylation signal for each fluorescent channel individually. Merged probes are labeled with [cpg_id]_MERG for the probe names. Merged probes' sesame detection p-values are calculated with the bates distribution.

7. **post_merge**: beta values from standard pipeline, but replicate probes are averaged *after beta values calculated*. Merged probes are labeled with [cpg_id]_MERG for the probe names. Merged probes' sesame detection p-values are calculated with the bates distribution.

8. **reference**: methylation percent processed from whole genome bisulfate sequencing data.

```{r processing, include = TRUE, cache = FALSE}

# Using a persistent app directory to cache output for this code chunk
user_data_path <- rappdirs::user_data_dir("tdhia")
data_filepath <- file.path(paste0(user_data_path, "/vignette_imprint_vs_wgbs_data.rds"))
dir.create(data_filepath, showWarnings = FALSE, recursive = TRUE)

if (!file.exists(data_filepath)) {
  # Store raw output (cpg beta matrix) of each pipeline (and gold standard reference) into a named list
  # to make downstream processing easier.
  raw_list <- list()
  
  # Pipeline 1: default                                                ###########
  #_______________________________________________________________________________
  # Standard base pipeline
  # 1) Load probe data and merge
  probe_beta <- 
    load_idata_to_probes(idat_dir_paths = idat_dir_path, enforce_req_idats = TRUE, 
                         merge_replicates = NULL, db_flag = TRUE)
  # 3) Filter probes results
  filt_probe_beta <-
    filter_probes(probe_beta = probe_beta, discard_unmapped_probes =TRUE ,
                  max_sig_pval = 0.2, max_probe_fail_rate = 0.25, db_flag = FALSE,
                  set_failed_betas_na = FALSE, discard_failed_probes = TRUE, min_design_score = .20)
  # 4) Merge probe measurement to cpg sites
  cpg_beta <- convert_probes_to_cpgs(filt_probe_beta, db_flag = TRUE)
  icr_beta <- convert_cpgs_to_icrs(cpg_beta)
  raw_list$default <- cpg_beta$cpg_beta_df
  
  
  # Pipeline 2: quant_norm                                                ###########
  #_______________________________________________________________________________
  # Standard pipeline, but performed quantile normalization (patient-wise) during 
  # cpg conversion step.
  #    Flags: convert_probes_to_cpgs:quantile_norm = TRUE
  cpg_beta <- convert_probes_to_cpgs(filt_probe_beta, db_flag = FALSE, quantile_norm = TRUE)
  raw_list$quant_norm <- cpg_beta$cpg_beta_df
  
  
  # Pipeline 3: cpg_smooth                                                ###########
  #_______________________________________________________________________________
  # Standard pipeline, but cpg smoothing is performed (moving average of beta values 
  # from adjacent sites within the same ICR) 
  #    Flags: convert_probes_to_cpgs:smooth_adj_cpgs = TRUE
  cpg_beta <- convert_probes_to_cpgs(filt_probe_beta, db_flag = FALSE, smooth_adj_cpgs = TRUE)
  raw_list$cpg_smooth <- cpg_beta$cpg_beta_df
  
  
  # Pipeline 4: oocyte support data                                    ###########
  #_______________________________________________________________________________
  # Standard pipeline, but only cpg's that have experiment-based oocyte support 
  # data are included (from 384 ICRs).
  # Load validation data
  df_valid <- as.data.frame(data.table::fread(paste0(system.file("data-raw", package = "tdhia"), "/validated_ICRs_11152024.bed")))
  colnames(df_valid) <- c("chr", "start", "finish", "icr")
  # determine which rows in manifest have validated cpgs
  is_validated <- sapply(1:nrow(mft), function(x) {any((
    mft[x,]$CHR==df_valid$chr) &
      (mft[x,]$MAPINFO > df_valid$start &
         mft[x,]$MAPINFO < df_valid$finish ))})
  validated_cpg_ids <- mft$Name[is_validated]
  # Subset of cpg sites with direct oocyte data supporting icr
  raw_list$oocyte <- cpg_beta$cpg_beta_df[rownames(cpg_beta$cpg_beta_df) %in%
                                            validated_cpg_ids,]
  
  
  # Pipeline 5: low_p_val                                             ############
  #_______________________________________________________________________________
  # Standard pipeline, but using a strict p-value threshold of 0.05 for sesame 
  # signal detection p-value instead of the usual
  #    Flags: max_sig_pval = 0.05
  # 3) Filter probes results
  filt_probe_beta <-
    filter_probes(probe_beta = probe_beta, discard_unmapped_probes =TRUE ,
                  max_sig_pval = 0.05, max_probe_fail_rate = 0.1, db_flag = FALSE,
                  set_failed_betas_na = FALSE, discard_failed_probes = TRUE, 
                  min_design_score = .20)
  # 4) Merge probe measurement to cpg sites
  cpg_beta <- convert_probes_to_cpgs(filt_probe_beta, db_flag = TRUE)
  raw_list$low_p_val <- cpg_beta$cpg_beta_df
  
  
  # Pipeline 6: design_thresh                                         ###########
  #_______________________________________________________________________________
  # Standard pipeline, excluding probes with low design score (0.30) are excluded.
  #    Flags: min_design_score = .30
  # 3) Filter probes results
  filt_probe_beta <-
    filter_probes(probe_beta = probe_beta, discard_unmapped_probes =TRUE ,
                  max_sig_pval = 0.2, max_probe_fail_rate = 0.25, db_flag = FALSE,
                  set_failed_betas_na = FALSE, discard_failed_probes = TRUE, min_design_score = .5)
  # 4) Merge probe measurement to cpg sites
  cpg_beta <- convert_probes_to_cpgs(filt_probe_beta, db_flag = FALSE)
  raw_list$design_thresh <- cpg_beta$cpg_beta_df
  
  
  # Pipeline 7: pre_merge                                               ###########
  #_______________________________________________________________________________
  #  Standard pipeline, but replicate probes are averaged by methylation and
  # unmethylation score for each fluorescent channel individually. Merged probes 
  # are labeled with [cpg_id]_MERG for the probe names. Merged probes' sesame 
  # detection p-values are calculated with the bates distribution. 
  #  There is a procedure that determine whether to merge probes or only keep one 
  # probe based on which has the smallest detection p-value.
  # 1) Load probe data and merge
  probe_beta <- 
    load_idata_to_probes(idat_dir_paths = idat_dir_path, enforce_req_idats = TRUE, 
                         merge_replicates = "pre_beta", db_flag = TRUE)
  # 3) Filter probes results
  filt_probe_beta <-
    filter_probes(probe_beta = probe_beta, discard_unmapped_probes =TRUE ,
                  max_sig_pval = 0.2, max_probe_fail_rate = 0.25, db_flag = FALSE,
                  set_failed_betas_na = FALSE, discard_failed_probes = TRUE, min_design_score = .20)
  # 4) Cpg matrix
  cpg_beta <- convert_probes_to_cpgs(filt_probe_beta, db_flag = FALSE)
  raw_list$pre_merge <- cpg_beta$cpg_beta_df
  
  
  # Pipeline 8: post_merge                                              ###########
  #_______________________________________________________________________________
  #  Standard pipeline, but replicate probes are averaged after beta value calculated. Merged 
  # probes are labeled with [cpg_id]_MERG for the probe names. Merged probes' sesame 
  # detection p-values are calculated with the bates distribution. 
  #  There is a procedure that determine whether to merge probes or only keep one 
  # probe based on which option has the smallest detection p-value (merged probe, 
  # probe #1, probe #2).
  # 1) Load probe data and merge
  probe_beta <- 
    load_idata_to_probes(idat_dir_paths = idat_dir_path, enforce_req_idats = TRUE, 
                         merge_replicates = "post_beta", db_flag = FALSE)
  # 3) Filter probes results
  filt_probe_beta <-
    filter_probes(probe_beta = probe_beta, discard_unmapped_probes =TRUE ,
                  max_sig_pval = 0.2, max_probe_fail_rate = 0.25, db_flag = FALSE,
                  set_failed_betas_na = FALSE, discard_failed_probes = TRUE, min_design_score = .20)
  # 4) Cpg matrix
  cpg_beta <- convert_probes_to_cpgs(filt_probe_beta, db_flag = FALSE)
  raw_list$post_merge <- cpg_beta$cpg_beta_df
  
  
  # Pipeline reference (wgbs)                                          ###########
  #_______________________________________________________________________________
  # Methylation ratio calculated for the same cpg sites using wgbs data.
  raw_list$reference <- tdhia::alzheimer_study_wbgs_shared_cpg
  save(raw_list, icr_beta, file = data_filepath)
} else {
  load(data_filepath)
}

```


## Visualize Distribution of beta values with and with quantile normalization

One of the options within the pipeline is to perform quantile normalization on beta values across patients. This makes normalizes the distribution across samples and is often used to compare data across methylation chips. Below is are densities plots of the beta values with (left) and without (right) normalization applied.

```{r, include = TRUE}

gg1 <- plot_beta_density(raw_list$default) + theme(legend.position = "none")

gg2 <- plot_beta_density(raw_list$quant_norm) + theme(legend.position = "none")
cowplot::plot_grid(gg1, gg2)

```

# Mean Correlation Imprintome Vs. WGBS

No we can compare the agreement of results between the imprintome array and WGBS. Beta values are first paired between the two technologies in a data frame that includes the union of all of the cpg sites found on the array. The WGBS dataset does not have coverage for all of these sites since it is a sequence based approach. On average the WGBS data has coverage for about 25% of the cpg sites found on the imprintome.

```{r, include = TRUE}

# Grab and sort data from each technology, paired by patient
df_grouped <- do.call(rbind, lapply(1:(length(names(raw_list))-1), function (x)
  group_by_mean(df_study = study_metadata, df_cpg_imp = raw_list[[x]],
                 df_cpg_wgbs = raw_list$reference, col_label = names(raw_list)[x],
                patient_id = x)))


```

## Comparison of Mean Pearson Correlation
To qualitatively visualize the agreement between these two technologies, we compare mean beta levels for the shared CPG sites between them. For the figure below, the beta values are calculated from a default pipeline and then averaged across all patients in a study group (black AD, black Control, white AD, white control).
```{r, include = TRUE}

# Show raw datapoints wgbs versus imprintome
gg <- ggplot2::ggplot(data = subset(df_grouped, pipeline_name == "default"), 
                      aes(x= wgbs, y = imp, color = pipeline_name)) +
  geom_point(size = 0.5) + theme_classic() + facet_wrap(vars(sample_id)) + 
  theme(legend.position ="none")
gg

```

## Comparison of Mean Pearson Correlation

We calculate the pearson correlation between the averaged beta measurements to a correlation coefficient for each of the study groups. Of the various imprintome pipelines tested, cpg smoothing and a design score threshold of > 0.5 increase the observed correlation coefficient with WGBS measurements.

```{r, include = TRUE}

# Calculate summary of correlation to gold standard
df_summary <- df_grouped %>% dplyr::group_by(sample_id, pipeline_name) %>% 
  dplyr::summarize(rho = cor(imp, wgbs, use = "complete.obs", method = "pearson"), 
                   .groups = "drop_last")

# Plot rho values for output from each version of the pipeline
gg <- ggplot2::ggplot(data = df_summary, aes(x= pipeline_name, y = rho, color = pipeline_name)) +
  geom_point() + theme_classic()
gg

```


## Per Patient Correlation of Imprintome vs WGBS

Now we can examine the correlation between beta values from the imprintome versus WGBS within each patient (instead of mean correlation for each study group). The pearson correlation for each patient is calculated (cpg sites that are not included in the WGBS data are removed). Once again, cpg smoothing and design score thresholding increase the observed mean correlation on a patient levl as well.

```{r, include = TRUE}

# Grab and sort data from each technology, paired by patient
df_grouped <- do.call(rbind, lapply(1:(length(names(raw_list))-1), 
       function(x) group_by_patient(df_study = pdata_wgbs_full, 
                                    df_cpg_imp = raw_list[[x]],   
                                    df_cpg_wgbs = raw_list$reference,
                                    col_label = names(raw_list)[x])))


# Calculate pearson rho between imprintome and wgbs for each patient
df_summary <- df_grouped %>% group_by(sample_id, name) %>% summarize(rho = cor(
  imp, wgbs, use = "complete.obs", method = "pearson"), .groups = "drop_last")
# df_summary <- summarize_patient_groups(df_grouped)


# Plot rho values for output from each version of the pipeline
gg <- ggplot2::ggplot(data = df_summary, aes(x= name, y = rho, color = name)) +
  geom_boxplot() + theme_classic()
gg


```

# Case-control analysis: Alzheimer Patient's 

Now we can perform a simple case control analysis on the data to see if there is a difference in mean beta values in healthy patients versus those with Alzheimers. Patients are separated between these two study groups,  and then a t-test is performed between the measurments from each group. P-values are corrected with a FDR correction base on the number of ICRs measured by the probes that passed quality control filtering.

```{r, include = TRUE}

case_control <- tdhia::analyze_case_control(raw_list$cpg_smooth, study_metadata$is_control, n_adjust = nrow(icr_beta$icr_beta_df))

# sum(case_control)

head(case_control[order(case_control$adj_pval, decreasing = FALSE),])

sum(case_control$adj_pval<0.05)

```

# Association Study: Imprintome Methylation Status and Alzheimer's

We can also perform an association study the also probes for differences in methylation between study groups, by also makes use of demographic information of the participants. We fit a series of linear models of the form:

disease_state ~ cpg_beta + Ethnicity + Age + Sex

1. **disease_state**: binary, TRUE is alzheimer's

2. **cpg_beta**: methylation beta value at a particular cpg site.

3. **Ethnicity**: factor with two levels: non-hispanic black and non-hispanic white.

4. **Age**: continuous variable, age of participant.

5. **Sex**: factor with 2 levels, male/ female.

Where disease_state is the response variable, and cpg_beta is the primary predictor, while also adjusting for ethnicity, age, and biological sex.



```{r, include = TRUE}

n.cores <- max(c(parallel::detectCores()-2,1))
df_alzheimer <- analyze_association(
  R = select(study_metadata, disease_state), P = as.data.frame(t(raw_list$cpg_smooth)),
  Pe = select(study_metadata, c(Ethnicity, Age, Sex)), family = "gaussian", 
  n_p_adj = nrow(icr_beta$icr_beta_df), db_flag = FALSE,rm.na.all = FALSE, 
  verbose = TRUE, impute_na = TRUE, max_p_val = 0.05, n.cores = n.cores)

```

## Significant ICR/CpG sites
```{r, include = TRUE}

# add ICR ID to the cpg sites
df_alzheimer$imp_site$icr_id <- match_cpg_id_to_icr_id(df_alzheimer$imp_site$Variable, AsNumeric = FALSE)
# Export result to simpler structure
df_res <- df_alzheimer$imp_site
# Add gene metadata to each cpg site
df_res_meta <- add_gene_metadata_to_icr(df_res)
# 
select(df_res_meta, c(icr_name, Variable, ADJ_P_VAL, is_icr_zinc, is_icr_highconf,
                      is_icr_medconf, is_icr_lowconf, Nearest.Transcript,
                      Distance.to.Nearest.Transcript))

```
