---
title: "Comparison of Methylation Measurements between Imprintome and WGBS"
author: "Bruce Corliss"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true # table of content true
    number_sections: true  ## if you want number sections at each table header
    highlight: tango  # specifies the syntax highlighting style
    theme: united 
    toc_float: true
    code_folding: hide
    toc_depth: 3  # upto three depths of headings (specified by #, ## and ###)
vignette: >
  %\VignetteIndexEntry{Comparison of Methylation Measurements between Imprintome and WGBS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Uncomment this next line if you want to knit this file on it's own, make sure working directory is set to tdhia repository base folder.
if (!("tdhia" %in% .packages())) devtools::load_all()

# detach(name = "package:tdhia", unload=TRUE)
# remove.packages("tdhia")
# tdhia_path <- devtools::build()
# install.packages(tdhia_path, lib='tdhia',repos = NULL, type="source",clean = TRUE)

```

# Introduction

This vignette will demonstrate how to load and process the imprintome files from the True Diagnostic Human Imprintome Array, along with conducting several statistical analysis on the data. We will use a dataset from:


> Carreras-Gallo N et al. Creation and Validation of the First Infinium DNA Methylation Array for the Human Imprintome. bioRxiv [Preprint]. 2024 Jan 16:2024.01.15.575646. doi: 10.1101/2024.01.15.575646. PMID: 38293193; PMCID: PMC10827131.


First, we will replicate and extend the results from Figure 4 of this paper. The dataset is comprised of 16 patients, half of which have varying degrees of Alzheimer's disease (N=8), and the other half are controls (N=8). Each patient has their imprintome methylation status measured from whole blood with whole genome bisulfite sequencing (WGBS) and from whole blood and the imprintome array. We will examine the agreement between these two technologies by comparing methylation beta values for each of the shared cpg site at the patient level, and also the mean methylation beta value across several patients in a study group.

Additionally, a case-control analysis will be performed with the imprintome dataset to test for changes in methylation between the Alzheimer patients and the control patients. A series of t-tests, one for each cpg site, will test for differences in the mean beta value between study groups.

Finally, we will perform an association study by using additional patient study data to probe for relationships between imprintome CpG methylation beta values and the presence of Alzheimer disease.


# Load packages and helper functions

We start by loading required packages for this vignette.  
1. **tdhia**: True Diagnostic Human Imprintome Array package.  
2. **rappdirs**: data storage path management.  
3. **tidyverse**: data processing and plotting.  
4. **cowplot**: plotting and saving plots to disk.  


```{r setup}

library(tdhia)
library(rappdirs)
library(dplyr)
library(tibble)
library(ggplot2)
library(cowplot)
library(data.table)
library(kableExtra)
library(ggVennDiagram)
```


We also load two user-defined helper functions for organizing data to for the comparative analysis.  
1. **group_by_patient()**: For each patient, group individual beta values from a specific imprintome pipeline, along with beta values from WGBS.  
2. **group_by_mean()**: For each study group, group mean beta values from a specific imprintome pipeline, along with beta values from WGBS.  


```{r setup2}

#' group_by_patient
#' @description For each patient, group beta/ methylation % values from a 
#' specific imprintome pipeline, along with methylation from WGBS.
#' 
#' @param df_study dataframe of study data, contains data for each patient to 
#' imprintome vs wgbs.
#' @param df_cpg_imp methylation data from a particular imprintome pipeline
#' @param df_cpg_wgbs methylation data from wgbs data
#' @param col_label string column label, used to label name of pipeline
#' 
#' @return
group_by_patient <- function(df_study, df_cpg_imp, df_cpg_wgbs, wgbs_counts, col_label, db_flag = TRUE) {
    if (db_flag) {save(list = ls(all.names = TRUE), file = "group_by_patient.RData")}
  # load(file = "group_by_patient.RData")
  
  df_list <- list()
  for (n in 1:nrow(df_study)) {
    # colnames(df_cpg_imp == df_study$Patient.ID[n]
    x = select(df_cpg_imp, one_of(df_study$Patient.ID[n])) %>% rename(imp = 1) %>%
      rownames_to_column("cpg_id")
    y = select(df_cpg_wgbs, one_of(df_study$wgbs_base_names[n])) %>%  rename(wgbs = 1) %>%
      rownames_to_column("cpg_id")
    z = select(wgbs_counts, one_of(df_study$wgbs_base_names[n])) %>%  rename(wgbs_counts = 1) %>%
      rownames_to_column("cpg_id")
    df_join0 <- left_join(x = x, y = y, by = "cpg_id", keep = FALSE, na_matches = "never",
                         unmatched= "drop", relationship = "one-to-one")
    df_join <- left_join(x = df_join0, y = z, by = "cpg_id", keep = FALSE, na_matches = "never",
                         unmatched= "drop", relationship = "one-to-one")
    
    # colnames(df_join) <- c("cpg_id", "imp","wgbs", "wgbs_count") 
    df_join$sample_id <- df_study$wgbs_base_names[n]
    df_join$treatment <- df_study$Treatment[n]
    df_join$pipeline_name <- col_label
    df_list[[n]] <- df_join
  }
  df_comp  <- do.call(rbind, df_list)
  return(df_comp)
}



#' group_by_mean
#' @description For study group, group mean beta/ methylation % values from a 
#' specific imprintome pipeline, along with methylation from WGBS.
#' 
#' @param df_study dataframe of study data, contains data for each patient to 
#' imprintome vs wgbs.
#' @param df_cpg_imp methylation data from a particular imprintome pipeline
#' @param df_cpg_wgbs methylation data from wgbs data
#' @param col_label string column label, used to label name of pipeline
#' 
#' @return
group_by_mean <- function(df_study, df_cpg_imp, df_cpg_wgbs, wgbs_counts, 
                          col_label, db_flag = TRUE, mean_max_na = 1) {
  if (db_flag) {save(list = ls(all.names = TRUE), file = "group_by_mean.RData")}
  # load(file = "vignettes/group_by_mean.RData")
  
  # Helper function
  nRowMeans <- function(x, max_na) {
    row_mean <- rowMeans(x, na.rm = TRUE); n_na <- rowMeans(is.na(x))
    row_mean[n_na > max_na] = NA; return(row_mean)
  }

  # Get unique study groups to average patients from each groups
  unq_study_group <- unique(df_study$study_group)
  
  df_list <- list()
  for (n in 1:length(unq_study_group)) {
    # Binary index of patients that match group
    bind <- df_study$study_group == unq_study_group[n]
    
    # Imprintome data, extract columns matching patients in group
    x <- data.frame(mean_imp = nRowMeans(select(df_cpg_imp, df_study$Patient.ID[bind]), max_na = mean_max_na)) %>%
      rownames_to_column("cpg_id")
    # WGBS data, extract columns matching patients in group
    y <- data.frame(mean_wgbs = nRowMeans(select(df_cpg_wgbs, df_study$wgbs_base_names[bind]),
                                          max_na = mean_max_na),
                    mean_wgbs_count = nRowMeans(select(wgbs_counts, df_study$wgbs_base_names[bind]), 
                                               max_na = mean_max_na)) %>% 
      rownames_to_column("cpg_id")
    
    df_join <- left_join(x = x, y = y, by = "cpg_id", keep = FALSE, na_matches = "never",
                         unmatched= "drop", relationship = "one-to-one")
    df_join$group_id <- unq_study_group[n]
    df_join$pipeline_name <- col_label
    df_join$patient_id <- paste(df_study$sampleID[which(bind)],collapse=",")
    df_list[[n]] <- df_join
  }
  df_comp  <- do.call(rbind, df_list)
  return(df_comp)
}

mft <- tdhia::manifest_v1A2_design_scores
```

# Load and clean study metadata

Study metadata is read from a dataframe included in the package, with columns defined below. 

0. *Rownames*: base file name for imprintome IDAT files.  
1. **ID**: patient index within this table (not study ID). 
2. **FileName**: filename of WGBS output files.  
3. **Treatment**: factor seperating patients by race and disease state (black control, 
white contro, black AD, white AD).  
4. **sampleID**: NEST or SHIP study ID (?). 
5. **X**: NEST or SHIP study ID (?).  
6. **Sample Chip**: identifier for each sample in this study.  
7. **Beadchip**: Index of True Diagnostic chip.  
8. **Row_Col**: location on chip.  
9. **is_control**: boolean, when true, denotes patinet is a healthy control patient.
10. **disease_state**: boolean, when true, denotes patients is a diseased patient with AD.  

```{r metadata, include = TRUE, cache = FALSE}
# Loading AD study data 
study_metadata <- tdhia::alzheimer_study_wbgs_shared_cpg_metadata

```


# Process Imprintome and WGBS data with Various Settings

There are various options within the TDHIA pipeline for processing imprintome data. We illustrate some of the primary options by processing the imprintome data with many deviations from the default settings for the pipeline. We can compare how well these different versions agree with results from WGBS data.  
1. **default**: beta values from standard default pipeline. The SeSame R package is used for processing the methylation data. Raw IDAT files are loaded, the Infinuium probe type is inferred, fluorescent signal is subtracted with Norm-Exp de-convolution that is parameterized with out-of-band probes, and a non-linear dye bias correction is applied. 
2. **epic_v2**: shared cpg sites with Epiv V2 Illumina methylation array.  
3. **quant_norm**: beta values from standard pipeline, but performed quantile normalization (patient-wise) during cpg conversion step.  
4. **cpg_smooth**: beta values from standard pipeline, but cpg smoothing is performed (moving average of beta values from adjacent sites within the same ICR).  
5. **gametic**: filter the CpG sites by those that have complete experimental gametic evidence of 50% methylation status.
6. **low_p_val**: beta values from standard pipeline, but using a strict p-value threshold of 0.05 for sesame signal detection p-value instead of the usual.  
7. **design_thresh**: exclude probes that have an illumina probe design score below 0.5.  
8. **pre_merge**: beta values from standard pipeline, but replicate probes are averaged *before beta values are calculated*, meaning the averaged by methylation and unmethylation signal for each fluorescent channel individually. Merged probes are labeled with [cpg_id]_MERG for the probe names. Merged probes' sesame detection p-values are calculated with the bates distribution.  
9. **post_merge**: beta values from standard pipeline, but replicate probes are averaged *after beta values calculated*. Merged probes are labeled with [cpg_id]_MERG for the probe names. Merged probes' sesame detection p-values are calculated with the bates distribution.  
10. **reference**: methylation percent processed from whole genome bisulfate sequencing data.  

```{r processing, include = TRUE, cache = FALSE}

# Using a persistent app directory to cache output for this code chunk
user_data_path <- rappdirs::user_data_dir("tdhia")
data_filepath <- file.path(paste0(user_data_path, "/vignette_imprint_vs_wgbs_data.rds"))
dir.create(dirname(data_filepath), showWarnings = FALSE, recursive = TRUE)
idat_dir_path <- system.file("extdata", package="tdhia")


# Get whitelist of CpG sites that have gametic validation
# df_valid <- as.data.frame(data.table::fread(paste0(system.file("data-raw", package = "tdhia"), "/validated_ICRs_11152024.bed")))
# colnames(df_valid) <- c("chr", "start", "finish", "icr")
df_valid <- tdhia::df_gametic_icrs
# determine which rows in manifest have validated cpgs
is_validated <- sapply(1:nrow(mft), function(x) {any((
  mft[x,]$CHR==df_valid$chr) & (mft[x,]$MAPINFO > df_valid$start &
       mft[x,]$MAPINFO < df_valid$finish ))})
gametic_cpg_ids <- mft$Name[is_validated]
# Number of gametic evidence ICRs
# length(unique(mft$icr_id[is_validated]))


if (!file.exists(data_filepath)) {
  # Store raw output (cpg beta matrix) of each pipeline (and gold standard reference) into a named list
  # to make downstream processing easier.
  pipelines_out <- list()
  
  # Pipeline 1: default                                                ###########
  #_______________________________________________________________________________
  # Standard base pipeline
  # A) Load probe data and merge
  probe_beta <- 
    load_idata_to_probes(idat_dir_paths = idat_dir_path, enforce_req_idats = TRUE, 
                         merge_replicates = NULL, db_flag = TRUE)
   filt_probe_beta <-
    filter_probes(probe_beta = probe_beta, discard_unmapped_probes =TRUE ,
                  max_sig_pval = 0.2, max_probe_fail_rate = 0.25, db_flag = FALSE,
                  set_failed_betas_na = FALSE, discard_failed_probes = TRUE,
                  min_design_score = .20, icr_conf_levels = NULL)
  cpg_beta <- convert_probes_to_cpgs(filt_probe_beta, db_flag = FALSE)
  pipelines_out$no_filter <- cpg_beta$cpg_beta_df
  
  # B) Filter probes results
  filt_probe_beta <-
    filter_probes(probe_beta = probe_beta, discard_unmapped_probes =TRUE ,
                  max_sig_pval = 0.2, max_probe_fail_rate = 0.25, db_flag = FALSE,
                  set_failed_betas_na = TRUE, discard_failed_probes = TRUE,
                  min_design_score = .20, icr_conf_levels = NULL)
  # C) Merge probe measurement to cpg sites
  cpg_beta <- convert_probes_to_cpgs(filt_probe_beta, db_flag = FALSE)
  icr_beta <- convert_cpgs_to_icrs(cpg_beta)
  pipelines_out$default <- cpg_beta$cpg_beta_df
  
  
  
  # Pipeline 2: epicv2                                                 #########
  #_______________________________________________________________________________
  # Standard pipeline, but only including cpg sites that are shared with Illumina 
  # epic version 2 array
  pipelines_out$epicv2 <- cpg_beta$cpg_beta_df[rownames(cpg_beta$cpg_beta_df) %in%
                                            tdhia::epicv2_shared_cpg_sites$Name,]
  
  
  # Pipeline 3: wgbs_count                                             #########
  #______________________________________________________________________________
  # Standard pipeline, but setting all measurements from imprintome are included 
  # if the corresponding wgbs count is greater than some threshold.
  cpg_beta_df2 = cpg_beta$cpg_beta_df
  # Remove any CpG beta value if it's corresponding count in wgbs is too low
  # must go patient-by-patient to assemble a blacklist of cpg sites with too low count
  
  for (n in 1:ncol(cpg_beta_df2)) {
    # CpGs with count < 5 or NA values
  cpg_black_list <- rownames(tdhia::alzheimer_study_wbgs_shared_cpg$count)[
    which(tdhia::alzheimer_study_wbgs_shared_cpg$count %>% select(all_of(n)) < 5 |
            tdhia::alzheimer_study_wbgs_shared_cpg$count %>% select(all_of(n)) %>% is.na())]
  cpg_beta_df2[rownames(cpg_beta_df2) %in% cpg_black_list, n] = NA
  }
  pipelines_out$wgbs_count  <- cpg_beta_df2
  
  # Pipeline 3: quant_norm                                                ###########
  #_______________________________________________________________________________
  # Standard pipeline, but performed quantile normalization (patient-wise) during 
  # cpg conversion step.
  #    Flags: convert_probes_to_cpgs:quantile_norm = TRUE
  cpg_beta <- convert_probes_to_cpgs(filt_probe_beta, db_flag = FALSE, quantile_norm = TRUE)
  pipelines_out$quant_norm <- cpg_beta$cpg_beta_df
  
  
  # Pipeline 4: cpg_smooth                                                ###########
  #_______________________________________________________________________________
  # Standard pipeline, but cpg smoothing is performed (moving average of beta values 
  # from adjacent sites within the same ICR) 
  #    Flags: convert_probes_to_cpgs:smooth_adj_cpgs = TRUE
  cpg_beta <- convert_probes_to_cpgs(filt_probe_beta, db_flag = FALSE, smooth_adj_cpgs = TRUE)
  pipelines_out$cpg_smooth <- cpg_beta$cpg_beta_df
  
  
  # Pipeline 5: gametic support data                                    ###########
  #_______________________________________________________________________________
  # Standard pipeline, but only cpg's that have experiment-based gametic support 
  # data are included (from 384 ICRs).
  # Subset of cpg sites with direct gametic data supporting icr
  pipelines_out$gametic <- cpg_beta$cpg_beta_df[rownames(cpg_beta$cpg_beta_df) %in%
                                            gametic_cpg_ids,]
  
  
  # Pipeline 6: low_p_val                                             ############
  #_______________________________________________________________________________
  # Standard pipeline, but using a strict p-value threshold of 0.05 for sesame 
  # signal detection p-value instead of the usual
  #    Flags: max_sig_pval = 0.05
  # C) Filter probes results
  filt_probe_beta <-
    filter_probes(probe_beta = probe_beta, discard_unmapped_probes =TRUE ,
                  max_sig_pval = 0.05, max_probe_fail_rate = 0.1, db_flag = FALSE,
                  set_failed_betas_na = FALSE, discard_failed_probes = TRUE, 
                  min_design_score = .20)
  # D) Merge probe measurement to cpg sites
  cpg_beta <- convert_probes_to_cpgs(filt_probe_beta, db_flag = TRUE)
  pipelines_out$low_p_val <- cpg_beta$cpg_beta_df
  
  
  # Pipeline 7: design_thresh                                         ###########
  #_______________________________________________________________________________
  # Standard pipeline, excluding probes with low design score (0.30) are excluded.
  #    Flags: min_design_score = .30
  # C) Filter probes results
  filt_probe_beta <-
    filter_probes(probe_beta = probe_beta, discard_unmapped_probes =TRUE ,
                  max_sig_pval = 0.2, max_probe_fail_rate = 0.25, db_flag = FALSE,
                  set_failed_betas_na = FALSE, discard_failed_probes = TRUE, min_design_score = .5)
  # D) Merge probe measurement to cpg sites
  cpg_beta <- convert_probes_to_cpgs(filt_probe_beta, db_flag = FALSE)
  pipelines_out$design_thresh <- cpg_beta$cpg_beta_df
  
  
  # Pipeline 8: pre_merge                                               ###########
  #_______________________________________________________________________________
  #  Standard pipeline, but replicate probes are averaged by methylation and
  # unmethylation score for each fluorescent channel individually. Merged probes 
  # are labeled with [cpg_id]_MERG for the probe names. Merged probes' sesame 
  # detection p-values are calculated with the bates distribution. 
  #  There is a procedure that determine whether to merge probes or only keep one 
  # probe based on which has the smallest detection p-value.
  # A) Load probe data and merge
  probe_beta <- 
    load_idata_to_probes(idat_dir_paths = idat_dir_path, enforce_req_idats = TRUE, 
                         merge_replicates = "pre_beta", db_flag = TRUE)
  # B) Filter probes results
  filt_probe_beta <-
    filter_probes(probe_beta = probe_beta, discard_unmapped_probes =TRUE ,
                  max_sig_pval = 0.2, max_probe_fail_rate = 0.25, db_flag = FALSE,
                  set_failed_betas_na = TRUE, discard_failed_probes = TRUE, min_design_score = .20)
  # C) Cpg matrix
  cpg_beta <- convert_probes_to_cpgs(filt_probe_beta, db_flag = FALSE)
  pipelines_out$pre_merge <- cpg_beta$cpg_beta_df
  
  
  # Pipeline 9: post_merge                                              ###########
  #_______________________________________________________________________________
  #  Standard pipeline, but replicate probes are averaged after beta value calculated. Merged 
  # probes are labeled with [cpg_id]_MERG for the probe names. Merged probes' sesame 
  # detection p-values are calculated with the bates distribution. 
  #  There is a procedure that determine whether to merge probes or only keep one 
  # probe based on which option has the smallest detection p-value (merged probe, 
  # probe #1, probe #2).
  # A) Load probe data and merge
  probe_beta <- 
    load_idata_to_probes(idat_dir_paths = idat_dir_path, enforce_req_idats = TRUE, 
                         merge_replicates = "post_beta", db_flag = FALSE)
  # B) Filter probes results
  filt_probe_beta <-
    filter_probes(probe_beta = probe_beta, discard_unmapped_probes =TRUE ,
                  max_sig_pval = 0.2, max_probe_fail_rate = 0.25, db_flag = FALSE,
                  set_failed_betas_na = FALSE, discard_failed_probes = TRUE, min_design_score = .20)
  # C) Cpg matrix
  cpg_beta <- convert_probes_to_cpgs(filt_probe_beta, db_flag = FALSE)
  pipelines_out$post_merge <- cpg_beta$cpg_beta_df
  
  
  # Pipeline reference (wgbs)                                          ###########
  #_______________________________________________________________________________
  # Methylation ratio calculated for the same cpg sites using wgbs data.
  pipelines_out$reference <- tdhia::alzheimer_study_wbgs_shared_cpg$ratio
  
  save(pipelines_out, icr_beta, file = data_filepath)
} else {
  load(data_filepath)
}



  
```


## Visualize Distribution of beta values with and with quantile normalization

One of the processing options within the pipeline is to perform quantile normalization on beta values across patients. This makes normalizes the distribution across samples and is often used to compare data across methylation chips. Below is are densities plots of the beta values with (left panel) and without (right panel) normalization applied.

```{r, include = TRUE, out.width="70%"}
gg1 <- plot_beta_density(pipelines_out$default) + theme(legend.position = "none")
gg2 <- plot_beta_density(pipelines_out$quant_norm) + theme(legend.position = "none")
cowplot::plot_grid(gg1, gg2)

```

# Mean Correlation Imprintome Vs. WGBS

Now we can compare the agreement of results between the imprintome array and WGBS. Beta values are first paired between the two technologies in a data frame that includes all of the cpg sites found on the array. The WGBS dataset does not have coverage for all of these sites since it is a sequence based approach. On average the WGBS data has coverage for about 25% of the cpg sites found on the imprintome.

```{r, include = TRUE}
# Arrange beta values for plotting, both imprintome and wgbs, paired by patient
df_grouped <- do.call(rbind, lapply(1:(length(names(pipelines_out))-1), function (x)
  group_by_mean(df_study = study_metadata, df_cpg_imp = pipelines_out[[x]],
                 df_cpg_wgbs = pipelines_out$reference, wgbs_counts = tdhia::alzheimer_study_wbgs_shared_cpg$count, 
                col_label = names(pipelines_out)[x], mean_max_na = 3)))

```

## Comparison of Mean Pearson Correlation

To visualize the agreement between these two technologies, we compare mean beta levels for the shared CPG sites between them. For the figure below, the beta values are calculated from a default pipeline and then averaged across all patients in a study group (black AD, black Control, white AD, white control).

```{r, include = TRUE, out.width="70%"}
# Show raw datapoints wgbs versus imprintome
gg <- ggplot2::ggplot(data = subset(df_grouped, pipeline_name == "default"), 
                      aes(x= mean_wgbs, y = mean_imp, color = mean_wgbs_count)) +
  geom_point(size = 0.5) + theme_classic() + facet_wrap(vars(group_id)) + 
  scale_color_gradient( guide = guide_colorbar(title = "WGBS Mean Count"), limits = c(0, 5))
gg

```

## Comparison of Mean Pearson Correlation {.tabset}

We calculate the pearson correlation of the averaged beta measurements (for each patient study group) between imprintome and WGBS measurements. The analysis is completed three times, with a different pipeline applied first before all other pipelines are applied:

1. **Base Pipeline**: basic imprintome pipeline.
2. **Gametic Evidence Default**: Only CpG sites with experimental evidence of 50% gametic methylation are included.
3. **Design Score Default**: only probes with a design score (> 0.5) is included in the output.  
  
Across all analysis, only CpG sites are included if they have no missing data across all 4 samples within the group for either imprintome or WGBS data.

### Default Pipeline

```{r, include = TRUE, out.width="80%", align="center"}

# Arrange beta values for plotting, both imprintome and wgbs, paired by patient
df_grouped <- do.call(rbind, lapply(1:(length(names(pipelines_out))-1), function (x)
  group_by_mean(df_study = study_metadata, df_cpg_imp = pipelines_out[[x]],
                 df_cpg_wgbs = pipelines_out$reference, wgbs_counts = tdhia::alzheimer_study_wbgs_shared_cpg$count, 
                col_label = names(pipelines_out)[x], mean_max_na = 0, db_flag = TRUE)))
# Calculate summary of correlation to gold standard
df_mean_summary <- df_grouped %>% dplyr::filter(!is.na(mean_imp) & !is.na(mean_wgbs)) %>%
  dplyr::group_by(group_id, pipeline_name) %>% 
  dplyr::summarize(rho = cor(mean_imp, mean_wgbs, use = "complete.obs", method = "pearson"), 
                   n = length(mean_imp),
                   .groups = "drop_last")

df_pipeline_summary <- df_mean_summary %>%  dplyr::group_by(pipeline_name) %>%  
  dplyr::summarize(min_rho = min(rho), max_rho = max(rho), mean_rho = mean(rho), mean_n = mean(n))

# Plot rho values for output from each version of the pipeline
gg1 <- ggplot2::ggplot(data = df_mean_summary, aes(x = pipeline_name, y = rho, color = group_id)) +
  geom_crossbar(data = df_pipeline_summary,
                aes(ymin = min_rho, y = mean_rho, ymax = max_rho), color = "grey50") +
  geom_point() + theme_classic() + coord_flip() + ylab("Pearson Rho") + ylim(c(0.4, 0.75)) +
  theme(legend.position = "right", axis.title.y=element_blank(), axis.text.y=element_blank())
# gg
gg2 <- ggplot2::ggplot(data = df_pipeline_summary, aes(x = pipeline_name, y = mean_n)) +
  geom_bar(stat = "identity") + theme_classic() + xlab("") + ylab("Shared CpG Count") + 
  coord_flip() + scale_y_continuous(breaks = scales::pretty_breaks(n = 3))
# gg2
ggf <- plot_grid(gg2, gg1, rel_widths = c(1, 2))
ggf

```

### Gametic Evidence Default
  
```{r, include = TRUE, out.width="80%", align="center"}

# Apply whitelist to all pipeline outputs and reference pipelines
pipelines_gametic <- lapply(pipelines_out, function(x) x[rownames(x) %in% gametic_cpg_ids,])
# Apply whitelist to wgbs count matrix also so both have same dimensions/ rows
wgbs_counts <- tdhia::alzheimer_study_wbgs_shared_cpg$count
wgbs_counts_ds <- wgbs_counts[rownames(wgbs_counts) %in% gametic_cpg_ids,] 
# all(rownames(wgbs_counts_ds) == rownames(pipelines_gametic$reference))

# Arrange beta values for plotting, both imprintome and wgbs, paired by patient
df_grouped <- do.call(rbind, lapply(1:(length(names(pipelines_gametic))-1), function (x)
  group_by_mean(df_study = study_metadata, df_cpg_imp = pipelines_gametic[[x]],
                 df_cpg_wgbs = pipelines_gametic$reference, 
                wgbs_counts = wgbs_counts_ds, mean_max_na = 0,
                col_label = names(pipelines_gametic)[x], db_flag = TRUE)))

# Calculate summary of correlation to gold standard
df_mean_summary <- df_grouped %>% dplyr::filter(!is.na(mean_imp) & !is.na(mean_wgbs)) %>%
  dplyr::group_by(group_id, pipeline_name) %>% 
  dplyr::summarize(rho = cor(mean_imp, mean_wgbs, use = "complete.obs", method = "pearson"), 
                   n = length(mean_imp),
                   .groups = "drop_last")

df_pipeline_summary <- df_mean_summary %>%  dplyr::group_by(pipeline_name) %>%  
  dplyr::summarize(min_rho = min(rho), max_rho = max(rho), mean_rho = mean(rho), mean_n = mean(n))

# Plot rho values for output from each version of the pipeline
gg1 <- ggplot2::ggplot(data = df_mean_summary, aes(x = pipeline_name, y = rho, color = group_id)) +
  geom_crossbar(data = df_pipeline_summary,
                aes(ymin = min_rho, y = mean_rho, ymax = max_rho), color = "grey50") +
  geom_point() + theme_classic() + coord_flip() + ylab("Pearson Rho") + ylim(c(0.4, 0.75)) +
  theme(legend.position = "right", axis.title.y=element_blank(), axis.text.y=element_blank())
# gg
gg2 <- ggplot2::ggplot(data = df_pipeline_summary, aes(x = pipeline_name, y = mean_n)) +
  geom_bar(stat = "identity") + theme_classic() + xlab("") + ylab("Shared CpG Count")  +
  coord_flip() + scale_y_continuous(breaks = scales::pretty_breaks(n = 3))
# gg2
ggf <- plot_grid(gg2, gg1, rel_widths = c(1, 2))
ggf

```


### Design Score Default

```{r, include = TRUE, out.width="80%", align="center"}

# Keep only CpG sites associated with high design score probes from imprintome 
mft <- tdhia::manifest_v1A2_design_scores
# Get white list of high design score cpg sites
cpg_white_list <- mft$Name[(mft$Design.Score >= 0.5) & !is.na(mft$Design.Score)]
# Apply whitelist to all pipeline outputs and reference pipelines
pipelines_design <- lapply(pipelines_out, function(x) x[rownames(x) %in% cpg_white_list,])
# Apply whitelist to wgbs count matrix also so both have same dimensions/ rows
wgbs_counts <- tdhia::alzheimer_study_wbgs_shared_cpg$count
wgbs_counts_ds <- wgbs_counts[rownames(wgbs_counts) %in% cpg_white_list,] 
# all(rownames(wgbs_counts_ds) == rownames(pipelines_design$reference))

# Arrange beta values for plotting, both imprintome and wgbs, paired by patient
df_grouped <- do.call(rbind, lapply(1:(length(names(pipelines_design))-1), function (x)
  group_by_mean(df_study = study_metadata, df_cpg_imp = pipelines_design[[x]],
                 df_cpg_wgbs = pipelines_design$reference, 
                wgbs_counts = wgbs_counts_ds, mean_max_na = 0,
                col_label = names(pipelines_design)[x], db_flag = TRUE)))

# Calculate summary of correlation to gold standard
df_mean_summary <- df_grouped %>% dplyr::filter(!is.na(mean_imp) & !is.na(mean_wgbs)) %>%
  dplyr::group_by(group_id, pipeline_name) %>% 
  dplyr::summarize(rho = cor(mean_imp, mean_wgbs, use = "complete.obs", method = "pearson"), 
                   n = length(mean_imp),
                   .groups = "drop_last")

df_pipeline_summary <- df_mean_summary %>%  dplyr::group_by(pipeline_name) %>%  
  dplyr::summarize(min_rho = min(rho), max_rho = max(rho), mean_rho = mean(rho), mean_n = mean(n))

# Plot rho values for output from each version of the pipeline
gg1 <- ggplot2::ggplot(data = df_mean_summary, aes(x = pipeline_name, y = rho, color = group_id)) +
  geom_crossbar(data = df_pipeline_summary,
                aes(ymin = min_rho, y = mean_rho, ymax = max_rho), color = "grey50") +
  geom_point() + theme_classic() + coord_flip() + ylab("Pearson Rho") + ylim(c(0.4, 0.75)) +
  theme(legend.position = "right", axis.title.y=element_blank(), axis.text.y=element_blank())
# gg
gg2 <- ggplot2::ggplot(data = df_pipeline_summary, aes(x = pipeline_name, y = mean_n)) +
  geom_bar(stat = "identity") + theme_classic() + xlab("") + ylab("Shared CpG Count")  + 
  coord_flip() + scale_y_continuous(breaks = scales::pretty_breaks(n = 3))
# gg2
ggf <- plot_grid(gg2, gg1, rel_widths = c(1, 2))
ggf

```


# Per Patient Correlation of Imprintome vs WGBS {.tabset}

Now we can examine the correlation between beta values within each patient for the imprintome versus WGBS measurements (instead of mean correlation for each study group). The pearson correlation for each patient is calculated (cpg sites that are not included in the WGBS or imprintome data are removed). THis analysis is repeated multiple times, with different setting applied to the imprintome and WGBS data.  

1. **Default**
2. **WGBS > 5**: CpG sites included if > 5 reads from WGBS output.  
3. **Gametic Evidence**: CpG sites with evidence of gametic origin of 50% methylation are included in the correlation comparison.  
4. **Imp DS > 0.5**: imprintome probes included with a design score > 0.5.  
  
## Default
  
```{r, include = TRUE, out.width="80%", align="center"}

# Grab and sort data from each technology, paired by patient
df_patients <- do.call(rbind, lapply(1:(length(names(pipelines_out))-1), 
       function(x) group_by_patient(df_study = study_metadata, 
                                    df_cpg_imp = pipelines_out[[x]],   
                                    df_cpg_wgbs = pipelines_out$reference,
                                    wgbs_counts = tdhia::alzheimer_study_wbgs_shared_cpg$count,
                                    col_label = names(pipelines_out)[x])))
df_patients$disease_state = ifelse(test = grepl(pattern = "control", x = df_patients$treatment),
                             yes= "control", no =  "ad")
# Calculate pearson rho between imprintome and wgbs for each patient
df_patients_rho <- df_patients %>% group_by(sample_id, pipeline_name, disease_state) %>% summarize(rho = cor(
  imp, wgbs, use = "complete.obs", method = "pearson"), .groups = "drop_last", 
  n_cpg = sum(!is.na(imp) & !is.na(wgbs))) %>% 
  rbind(df_patients %>% group_by(sample_id, pipeline_name) %>% summarize(disease_state = "all", rho = cor(
  imp, wgbs, use = "complete.obs", method = "pearson"), .groups = "drop_last", 
  n_cpg = sum(!is.na(imp) & !is.na(wgbs))))
df_patients_rho$disease_state <- 
  factor(df_patients_rho$disease_state, levels = c("control", "ad", "all"), ordered = TRUE)

# Counts of probes that pass QC for each method
df_patients_rho_summary <- df_patients_rho %>% group_by(pipeline_name) %>% summarize(mean_ncpg = mean(n_cpg))
# Plot rho values for output from each version of the pipeline
gg1 <- ggplot2::ggplot(data = df_patients_rho, aes(x= pipeline_name, y = rho)) +
  geom_boxplot() + facet_grid(cols = vars(disease_state)) + theme_classic() + coord_flip() +
  ylab("Pearson Rho") + ylim(c(0.3,0.8)) +
  theme( axis.title.y=element_blank(), axis.text.y=element_blank())
gg2 <- ggplot2::ggplot(data = df_patients_rho_summary, aes(x = pipeline_name, y = mean_ncpg)) +
   geom_bar(stat = "identity") + theme_classic() +
   xlab("Pipeline") + ylab("Shared CpG Count") +
   coord_flip()
ggf <- plot_grid(gg2, gg1, rel_widths = c(1, 1.6))
ggf

```
  
Figure: Per-patient correlation between Imprintome and WGBS. WGBS CpG sites included regardless of # of reads. (A) Bar chart of number of Imprintome probes for each pipeline. Pearson rho correlation coefficients for each pipeline (N = 16 patients).

## WGBS > 5 
  
```{r, include = TRUE, out.width="80%", align="center"}
# Apply a count threshold to the wgbs output beta values (reference)
#_______________________________________________________________________________
pipelines_wgbs_thresh <- pipelines_out
pipelines_wgbs_thresh$reference[tdhia::alzheimer_study_wbgs_shared_cpg$count<5]=NA

# Grab and sort data from each technology, paired by patient
df_patients <- do.call(rbind, lapply(1:(length(names(pipelines_wgbs_thresh))-1), 
       function(x) group_by_patient(df_study = study_metadata, 
                                    df_cpg_imp = pipelines_wgbs_thresh[[x]],   
                                    df_cpg_wgbs = pipelines_wgbs_thresh$reference,
                                    wgbs_counts = tdhia::alzheimer_study_wbgs_shared_cpg$count,
                                    col_label = names(pipelines_wgbs_thresh)[x])))
df_patients$disease_state = ifelse(test = grepl(pattern = "control", x = df_patients$treatment),
                             yes= "control", no =  "ad")
# Calculate pearson rho between imprintome and wgbs for each patient
df_patients_rho <- df_patients %>% group_by(sample_id, pipeline_name, disease_state) %>% summarize(rho = cor(
  imp, wgbs, use = "complete.obs", method = "pearson"), .groups = "drop_last", 
  n_cpg = sum(!is.na(imp) & !is.na(wgbs))) %>% 
  rbind(df_patients %>% group_by(sample_id, pipeline_name) %>% summarize(disease_state = "all", rho = cor(
  imp, wgbs, use = "complete.obs", method = "pearson"), .groups = "drop_last", 
  n_cpg = sum(!is.na(imp) & !is.na(wgbs))))
df_patients_rho$disease_state <- 
  factor(df_patients_rho$disease_state, levels = c("control", "ad", "all"), ordered = TRUE)

# Counts of probes that pass QC for each method
df_patients_rho_summary <- df_patients_rho %>% group_by(pipeline_name) %>% summarize(mean_ncpg = mean(n_cpg))
# Plot rho values for output from each version of the pipeline
gg1 <- ggplot2::ggplot(data = df_patients_rho, aes(x= pipeline_name, y = rho)) +
  geom_boxplot() + facet_grid(cols = vars(disease_state)) + theme_classic() + 
  coord_flip() + ylab("Pearson Rho") + ylim(c(0.3,0.8)) +
  theme( axis.title.y=element_blank(), axis.text.y=element_blank())
# gg1
gg2 <- ggplot2::ggplot(data = df_patients_rho_summary, aes(x = pipeline_name, y = mean_ncpg)) +
   geom_bar(stat = "identity") + theme_classic() +
   xlab("Pipeline") + ylab("Shared CpG Count") +
   coord_flip()
# gg2
ggf <- plot_grid(gg2, gg1, rel_widths = c(1, 1.6))
ggf

```
  
Figure: Per-patient correlation between Imprintome and WGBS. WGBS CpG sites filtered to those with > 5 reads. (A) Bar chart of number of Imprintome probes for each pipeline. Pearson rho correlation coefficients for each pipeline (N = 16 patients).
  
  
## WGBS > 5, Gametic Evidence

```{r, include = TRUE, out.width="80%", align="center"}
# Apply a count threshold to the wgbs output beta values (reference)
#_______________________________________________________________________________
pipelines_wgbs_thresh <- pipelines_out
pipelines_wgbs_thresh$reference[tdhia::alzheimer_study_wbgs_shared_cpg$count<5]=NA


# Apply whitelist to all pipeline outputs and reference pipelines, and wgbs count data
pipelines_design <- lapply(pipelines_wgbs_thresh, function(x) x[rownames(x) %in% gametic_cpg_ids,])
wgbs_counts <- tdhia::alzheimer_study_wbgs_shared_cpg$count
wgbs_counts_ds <- wgbs_counts[rownames(wgbs_counts) %in% gametic_cpg_ids,] 
# all(rownames(wgbs_counts_ds) == rownames(pipelines_design$reference))


# Grab and sort data from each technology, paired by patient
df_patients <- do.call(rbind, lapply(1:(length(names(pipelines_design))-1), 
       function(x) group_by_patient(df_study = study_metadata, 
                                    df_cpg_imp = pipelines_design[[x]],   
                                    df_cpg_wgbs = pipelines_design$reference,
                                    wgbs_counts = tdhia::alzheimer_study_wbgs_shared_cpg$count,
                                    col_label = names(pipelines_design)[x])))
df_patients$disease_state = ifelse(test = grepl(pattern = "control", x = df_patients$treatment),
                             yes= "control", no =  "ad")
# Calculate pearson rho between imprintome and wgbs for each patient
df_patients_rho <- df_patients %>% group_by(sample_id, pipeline_name, disease_state) %>% summarize(rho = cor(
  imp, wgbs, use = "complete.obs", method = "pearson"), .groups = "drop_last", 
  n_cpg = sum(!is.na(imp) & !is.na(wgbs))) %>% 
  rbind(df_patients %>% group_by(sample_id, pipeline_name) %>% summarize(disease_state = "all", rho = cor(
  imp, wgbs, use = "complete.obs", method = "pearson"), .groups = "drop_last", 
  n_cpg = sum(!is.na(imp) & !is.na(wgbs))))
df_patients_rho$disease_state <- 
  factor(df_patients_rho$disease_state, levels = c("control", "ad", "all"), ordered = TRUE)

# Counts of probes that pass QC for each method
df_patients_rho_summary <- df_patients_rho %>% group_by(pipeline_name) %>% summarize(mean_ncpg = mean(n_cpg))
# Plot rho values for output from each version of the pipeline
gg1 <- ggplot2::ggplot(data = df_patients_rho, aes(x= pipeline_name, y = rho)) +
  geom_boxplot()  + facet_grid(cols = vars(disease_state)) + theme_classic() + 
  coord_flip() + ylab("Pearson Rho") + ylim(c(0.3,0.8)) +
  theme( axis.title.y=element_blank(), axis.text.y=element_blank())
gg2 <- ggplot2::ggplot(data = df_patients_rho_summary, aes(x = pipeline_name, y = mean_ncpg)) +
   geom_bar(stat = "identity") + theme_classic() +
   xlab("Pipeline") + ylab("Shared CpG Count") +
   coord_flip()
ggf <- plot_grid(gg2, gg1, rel_widths = c(1, 1.6))
ggf

```
   
Figure: Per-patient correlation between Imprintome and WGBS. WGBS CpG sites filtered to those with > 5 reads and gametic origin of methylation. (A) Bar chart of number of Imprintome probes for each pipeline. Pearson rho correlation coefficients for each pipeline (N = 16 patients).
  
  
## WGBS > 5, Imp DS > 0.5 
  
```{r, include = TRUE, out.width="80%", align="center"}
# Apply a count threshold to the wgbs output beta values (reference)
#_______________________________________________________________________________
pipelines_wgbs_thresh <- pipelines_out
pipelines_wgbs_thresh$reference[tdhia::alzheimer_study_wbgs_shared_cpg$count<5]=NA

# Get white list of high design score cpg sites
cpg_white_list <- mft$Name[(mft$Design.Score >= 0.5) & !is.na(mft$Design.Score)]
# Apply whitelist to all pipeline outputs, reference pipelines, and wgbs count data
pipelines_design <- lapply(pipelines_wgbs_thresh, function(x) x[rownames(x) %in% cpg_white_list,])
wgbs_counts <- tdhia::alzheimer_study_wbgs_shared_cpg$count
wgbs_counts_ds <- wgbs_counts[rownames(wgbs_counts) %in% cpg_white_list,] 
# all(rownames(wgbs_counts_ds) == rownames(pipelines_design$reference))

# Grab and sort data from each technology, paired by patient
df_patients <- do.call(rbind, lapply(1:(length(names(pipelines_design))-1), 
       function(x) group_by_patient(df_study = study_metadata, 
                                    df_cpg_imp = pipelines_design[[x]],   
                                    df_cpg_wgbs = pipelines_design$reference,
                                    wgbs_counts = tdhia::alzheimer_study_wbgs_shared_cpg$count,
                                    col_label = names(pipelines_design)[x])))
df_patients$disease_state = ifelse(test = grepl(pattern = "control", x = df_patients$treatment),
                             yes= "control", no =  "ad")
# Calculate pearson rho between imprintome and wgbs for each patient
df_patients_rho <- df_patients %>% group_by(sample_id, pipeline_name, disease_state) %>% summarize(rho = cor(
  imp, wgbs, use = "complete.obs", method = "pearson"), .groups = "drop_last", 
  n_cpg = sum(!is.na(imp) & !is.na(wgbs))) %>% 
  rbind(df_patients %>% group_by(sample_id, pipeline_name) %>% summarize(disease_state = "all", rho = cor(
  imp, wgbs, use = "complete.obs", method = "pearson"), .groups = "drop_last", 
  n_cpg = sum(!is.na(imp) & !is.na(wgbs))))
df_patients_rho$disease_state <- 
  factor(df_patients_rho$disease_state, levels = c("control", "ad", "all"), ordered = TRUE)

# Counts of probes that pass QC for each method
df_patients_rho_summary <- df_patients_rho %>% group_by(pipeline_name) %>% summarize(mean_ncpg = mean(n_cpg))
# Plot rho values for output from each version of the pipeline
gg1 <- ggplot2::ggplot(data = df_patients_rho, aes(x= pipeline_name, y = rho)) +
  geom_boxplot()  + facet_grid(cols = vars(disease_state)) + theme_classic() +
  coord_flip() + ylab("Pearson Rho") + ylim(c(0.3,0.8)) +
  theme( axis.title.y=element_blank(), axis.text.y=element_blank())
gg2 <- ggplot2::ggplot(data = df_patients_rho_summary, aes(x = pipeline_name, y = mean_ncpg)) +
   geom_bar(stat = "identity") + theme_classic() +
   xlab("Pipeline") + ylab("Shared CpG Count") +
   coord_flip()
ggf <- plot_grid(gg2, gg1, rel_widths = c(1, 1.6))
ggf

```

Figure: Per-patient correlation between Imprintome and WGBS. WGBS CpG sites filtered to those with > 5 reads and only high quality probes probes included from imprintome data (design score > 0.5). (A) Bar chart of number of Imprintome probes for each pipeline. Pearson rho correlation coefficients for each pipeline (N = 16 patients).



# Case-control analysis: Alzheimer Patient's 

Now we can perform a simple case control analysis on the data to see if there is a difference in mean beta values in healthy patients versus those with Alzheimers. Patients are separated between these two study groups,  and then a t-test is performed between the measurments from each group. P-values are corrected with a FDR correction base on the number of ICRs measured by the probes that passed quality control filtering.

```{r, include = TRUE}

case_control <- tdhia::analyze_case_control(pipelines_out$pre_merge, study_metadata$is_control, n_adjust = nrow(icr_beta$icr_beta_df))

# sum(case_control)

head(case_control[order(case_control$adj_pval, decreasing = FALSE),])

sum(case_control$adj_pval<0.05)

```

# Association Study: Imprintome Methylation Status and Alzheimer's {.tabset}

We can also perform an association study the also probes for differences in methylation between study groups, by also makes use of demographic information of the participants. We fit a series of binomial models that include all participants, and then partition them between Black and White patients:

**Model 1**: disease_state ~ cpg_beta + Age + Sex, All patients  
**Model 2**: disease_state ~ cpg_beta + Age + Sex, Black patients  
**Model 3**: disease_state ~ cpg_beta + Age + Sex, White patients  

1. **disease_state**: binary, TRUE is Alzheimer's.  
2. **cpg_beta**: methylation beta value at a particular cpg site.  
3. **Age**: continuous variable, age of participant.  
4. **Sex**: factor with 2 levels, male/ female.  

Where disease_state is the response variable, and cpg_beta is the primary predictor, while also adjusting for ethnicity, age, and biological sex.



```{r association_study, include = TRUE}
n.cores <- max(c(parallel::detectCores()-2,1))

# disease_state ~ cpg + Age + Sex, All patients
df_all <- analyze_association(
  R = select(study_metadata, disease_state), P = as.data.frame(t(pipelines_out$default)),
  Pe = select(study_metadata, c(Age, Sex)), family = "binomial", 
  n_p_adj = nrow(icr_beta$icr_beta_df), db_flag = FALSE, rm.na.all = FALSE, 
  verbose = TRUE, impute_na = TRUE, max_p_val = 0.05, n.cores = n.cores)

# disease_state ~ cpg + Age + Sex, White only
ix_white <- study_metadata$Ethnicity == "NHW"
df_white <- analyze_association(
  R = select(study_metadata[ix_white,], disease_state), 
  P = as.data.frame(t(select(pipelines_out$pre_merge, study_metadata$Patient.ID[ix_white]))),
  Pe = select(study_metadata[ix_white,], c(Age, Sex)), family = "binomial", 
  n_p_adj = nrow(icr_beta$icr_beta_df), db_flag = FALSE,rm.na.all = FALSE, 
  verbose = TRUE, impute_na = TRUE, max_p_val = 0.05, n.cores = n.cores)

# disease_state ~ cpg + Age + Sex, Black patients
ix_black <- study_metadata$Ethnicity == "NHB"
df_black <- analyze_association(
  R = select(study_metadata[ix_black,], disease_state), 
  P = as.data.frame(t(select(pipelines_out$pre_merge, study_metadata$Patient.ID[ix_black]))),
  Pe = select(study_metadata[ix_black,], c(Age, Sex)), family = "binomial", 
  n_p_adj = nrow(icr_beta$icr_beta_df), db_flag = FALSE,rm.na.all = FALSE, 
  verbose = TRUE, impute_na = TRUE, max_p_val = 0.05, n.cores = n.cores)


```
  
**Significant CpG and ICR sites**
  
Now we can retrieve additional information for any CpG sites that are found to have an association, such as proximity to a zinc finger finger binding site, experimental evidence for it's associated ICR, and the nearest gene transcript.

```{r association_sig_icr, out.width = "50%", align = "center"}


df_result <- rbind(df_all$imp_site %>% mutate(Group = "All"), 
                   df_black$imp_site %>% mutate(Group = "Black"),
                   df_white$imp_site %>% mutate(Group = "White")) %>% 
  filter(ADJ_P_VAL < 0.05)
# add ICR ID to the cpg sites
df_result$icr_id <- match_cpg_id_to_icr_id(df_result$Variable, AsNumeric = FALSE)

# # Add gene metadata to each cpg site
df_result_meta <- add_gene_metadata_to_icr(df_result)
# # Output a subset of the table
df_display <- dplyr::select(
  df_result_meta, c(Group, icr_name, Variable, ADJ_P_VAL, is_icr_zinc, 
                    is_icr_highconf, is_icr_medconf, is_icr_lowconf, Nearest.Transcript,
                    Distance.to.Nearest.Transcript))

colnames(df_display) <- c("Group", "ICR", "CpG Site", "Adj p-val", "Zinc", "High", 
                          "Med", "Low", "Nearest.Transcript", "Distance")

disp_table_all = filter(df_display, Group == "All") %>% select(-Group)
disp_table_black = filter(df_display, Group == "Black") %>% select(-Group)
disp_table_white = filter(df_display, Group == "White") %>% select(-Group)


# Produce venn diagram of shared ICRs between groups
ggvenn <- venn_shared_icrs(df_imp_sig = df_result, colname = "Group", group_names = NULL, 
                             output_dir_path = NULL, name_suffix = "")

ggvenn$figure

```
  
Figure: venn diagram of the number of shared significant ICR sites between ethnicity groups from association study.

## All
  
Significant CpGs: `r sprintf("%.0f\n", nrow(disp_table_all))`.  
Significant ICRs:  `r sprintf(" %.0f\n", length(unique(disp_table_all$ICR)))`.
  
```{r, include = TRUE, echo = TRUE}
knitr::kable(disp_table_all, format = "html") %>% kableExtra::kable_styling()
```


## Black
  
Significant CpGs: `r sprintf("%.0f\n", nrow(disp_table_black))`.  
Significant ICRs:  `r sprintf(" %.0f\n", length(unique(disp_table_black$ICR)))`.
  
```{r, include = TRUE, echo = TRUE}
knitr::kable(disp_table_black, format = "html") %>% kableExtra::kable_styling()
```

## White
  
Significant CpGs: `r sprintf("%.0f\n", nrow(disp_table_white))`.  
Significant ICRs:  `r sprintf(" %.0f\n", length(unique(disp_table_white$ICR)))`.
  
```{r, include = TRUE, echo = TRUE}
knitr::kable(disp_table_white, format = "html") %>% kableExtra::kable_styling()
```

