---
title: "Analyzing TDHIA IDAT files"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analyzing TDHIA IDAT files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")

library(dplyr)
# here::i_am("README.md")
devtools::load_all()
```

Here we will demonstrate how to process IDAT files from the True Diagnostic Human
Imprintome Array (TDHIA) with the tdhia package.

## Dataset generation

We begin by first generating simulated data. Sesame imports IDAT files into a SigSet data object. We are going to artificially generate SigSet files. Let's do a dataset of 10 SigSets (IDATS).


```{r simulate}
n_patients = 10
ss <- tdhia::simulate_idat_data(n_patients)
```

We can now load this data into tdhia and calculate the beta values and signal 
detection p-values. This is returned in a named list, with the following fields:  
**probe_beta_df**: dataframe of beta values, probe_id x sample_id.  
**probe_pval_df**: dataframe of signal p-values, probe_id x sample_id.  
**platform**: string that describes platform for methylation array.  
**manifest**: dataframe of the manifest file used for SeSame processing.  
  


```{r probes}
probe_beta <-
    tdhia::load_idata_to_probes(idat_dir_paths = ss, multicore = 2, 
                         quantile_norm = FALSE, db_flag = TRUE)
```
  
You can access the probe beta matrix with the **probe_beta_df** field.
```{r visualize}
# Access probe beta matrix
head(probe_beta$probe_beta_df)
```


We can filter out low quality probes with the filt_probe_beta(). This function 
returns a named list with the same format as *probe_beta* above.  


```{r filter}
# 2) Filter probes that are not mapped and discard poor signal
# Output: rows: probe_id     x     columns: patient
filt_probe_beta <-
  tdhia::filter_probes(probe_beta = probe_beta, discard_unmapped_probes = TRUE ,
                max_sig_pval = 0.2, set_failed_betas_na = FALSE, 
                max_probe_fail_rate = 0.25,
                discard_failed_probes = TRUE, db_flag = FALSE)
```

You can access the filtered probe beta matrix with the **probe_beta_df** field.
```{r filter_visualize}
# Access filtered probe beta matrix
head(filt_probe_beta$probe_beta_df)
```



Once probe data has been filtered convert the probe beta matrix to a cpg data 
matrix, stored in a named list of the form:  
**cpg_beta_df**: dataframe of beta values, probe_id x sample_id.  
**platform**: string that describes platform for methylation array.  
**manifest**: dataframe of the manifest file used for SeSame processing.  
**cpg_n_probes**: number of probes for each cpg site.
  
Probe beta values that belong to the same cpg site for a given patient are averaged.  

```{r cpg}
#3  Convert probe beta matrix to a cpg beta matrix
# Output: rows: cpg_id     x     columns: patient
cpg_beta <- tdhia::convert_probes_to_cpgs(filt_probe_beta, quantile_norm = FALSE, 
                                   db_flag = TRUE, smooth_adj_cpgs = TRUE, multicore = 2)
```

You can access the filtered probe beta matrix with the **cpg_beta_df** field.
```{r cpg_visualize}
# Access filtered probe beta matrix
head(cpg_beta$cpg_beta_df)
```



Finally, we can convert the cpg beta matric to an icr beta matric, which is 
returned in a named list of the form:  
**icr_beta_df**: dataframe of beta values, icr_id x sample_id.  
**platform**: string that describes platform for methylation array.  
**manifest**: dataframe of the manifest file used for SeSame processing.  
**n_CpGs**: number of cpg sites in each icr site.
  
Probe beta values that belong to the same cpg site for a given patient are averaged.  


```{r icr}

#4  Convert probe beta matrix to an icr beta matrix
# Output: rows: icr_id     x     columns: patient
icr_beta <- tdhia::convert_cpgs_to_icrs(cpg_beta, max_icr_fail_rate = 0.2)

```

You can access the filtered probe beta matrix with the **probe_beta_df** field.
```{r}
# Access filtered probe beta matrix
head(icr_beta$icr_beta_df)
```

Let's simulate a case control analysis, where we will test for differences in CpG methylation state between a control group and an treatment group.

```{r}

groud_id = c(rep(0,floor(n_patients/2)), rep(1,n_patients - floor(n_patients/2)))

# Test for significant cpgs, corrected for number of ICRs
df_res <- analyze_case_control(cpg_beta$cpg_beta_df, groud_id, n_adjust = nrow(icr_beta$icr_beta_df))
sum(df_res$adj_sig)

# Test for significant ICRs
df_res <- analyze_case_control(icr_beta$icr_beta_df, groud_id)
sum(df_res$adj_sig)

```

Now we can simulate a associate study, where we check if there is a correlation 
between cpg methylation state and other patient variables.

First, a data frame is created with patient information, including any response 
variables or predictors. For actual analysis, this data is usually imported from
a csv file or similiar format.

```{r}

df_study = data.frame(id = 1:n_patients)
df_study$race <- as.factor(sample(c(0,1,2), 10, replace = TRUE))
df_study$exposure<- runif(n = n_patients, min = 0, max= 10)
df_study$smoked <- as.factor(sample(c(0,1), 10, replace = TRUE))
df_study

rownames(df_study) <- colnames(cpg_beta$cpg_beta_df)
```





```{r}

n.cores <- max(c(parallel::detectCores()-2,1))

df_complete <- analyze_association(R = select(df_study, exposure),
                                   P = as.data.frame(t(cpg_beta$cpg_beta_df)),
                                   Pe = select(df_study, -exposure),
                                   family = "gaussian", n_p_adj = nrow(icr_beta$icr_beta_df), db_flag = FALSE,
                                   rm.na.all = FALSE, verbose = TRUE, impute_na = TRUE,
                                   max_p_val = 0.05, n.cores = n.cores)

```
