---
title: "Analyzing TDHIA IDAT files"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
devtools::load_all()
```

Here we will demonstrate how to process IDAT files from the True Diagnostic Human
Imprintome Array (TDHIA) with the tdhia package.

## Dataset generation

We begin by first generating simulated data. Sesame imports IDAT files into a SigSet data object. We are going to artificially generate SigSet files. Let's do a dataset of 25 SigSets (IDATS).


```{r}
ss <- simulate_idat_data(25)
```

We can now load this data into tdhia and calculate the beta values and signal detection p-values.


```{r}

probe_beta <-
    load_idata_to_probes(idat_dir_paths = ss, multicore = TRUE, 
                         quantile_norm = FALSE)

```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
