---
title: "Introduction to TDHIA R Package"
author: "Bruce Corliss"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true # table of content true
    number_sections: true  ## if you want number sections at each table header
    highlight: tango  # specifies the syntax highlighting style
    theme: united 
    toc_float: true
    code_folding: hide
    toc_depth: 3  # upto three depths of headings (specified by #, ## and ###)
---


```{r setup, include = FALSE, purl = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

if (!("tdhia" %in% .packages())) devtools::load_all()

library(dplyr)

```

Here we will demonstrate how to process IDAT files from the True Diagnostic Human
Imprintome Array (TDHIA) with the tdhia package.


# Dataset generation

We will use a small dataset of imprintome methylation samples obtained from brain tissue, split between patients with and with Alzheimer disease. The data will be downloaded from [GSE268075]https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE268075. We can use a built-in GEO downloader in tdhia which will cache the data in a user directory associated with the package.

Some additional patient metadata are also added for us to complete this analysis.

```{r load dataset, eval=TRUE}


# Download dataset and cache from GEO
local_download <- tdhia::download_geo_methylation("GSE268075", quiet = F)
# Correct IDAT filenames to match SeSame assumed importation convention
idat_renames <- tdhia::correct_idat_names(file.path(local_download$directory, "idat"), rename = T)


# Load metadata from GEO
study_metadata <- read.csv(file.path(local_download$metadata_files))

# Define 
study_metadata <- study_metadata %>% dplyr::select(geo_accession, data_row_count, race.ch1, tissue.ch1, disease.ch1, supplementary_file)
study_metadata$Patient.ID <- study_metadata$supplementary_file %>%
  basename() %>% stringr::str_replace("_grn.IDAT.gz", "") 
study_metadata <- study_metadata %>% select(-supplementary_file)
study_metadata <- study_metadata %>% rename(race = race.ch1, tissue = tissue.ch1,  disease =  disease.ch1)
study_metadata$race = recode(study_metadata$race, "black" = "NHB", "white" = "NHW")

# Loading additional AD study data 
study_metadata2 <- tdhia::alzheimer_study_wbgs_shared_cpg_metadata
patient_id_geo_lookup  <- setNames(study_metadata$Patient.ID, study_metadata$Patient.ID %>%
   sub(pattern="GSM[0-9]*_", replacement = ""))
study_metadata2$Patient.ID <- patient_id_geo_lookup[study_metadata2$Patient.ID]

# Joining tables together
study_metadata <- study_metadata %>% left_join(study_metadata2 %>% select(
  Patient.ID, Ethnicity, Age, Sex, Brain.Region, Time, AD.Neuropathologic.changes, 
  wgbs_base_names), by = join_by(Patient.ID) )

# Add patient_ids with GEO numbers as IDs for patients across datasets
rownames(study_metadata) <- study_metadata$Patient.ID

# Custom variables for convenience
study_metadata$is_control = study_metadata$disease == "Control"
study_metadata$disease_state = study_metadata$disease == "Alzheimer Disease"
study_metadata$study_group = paste0(
  ifelse(study_metadata$race=="NHB", "Black", "White"), " " ,
  ifelse(study_metadata$disease=="Control", "Ctrl", "AD"))

# Remove incomplete rows
study_metadata <- filter(study_metadata, !is.na(wgbs_base_names))
 
# Remove extraneous variables
study_metadata <- study_metadata %>% select(-c(data_row_count, Brain.Region, Time))


head(study_metadata)
```


We specified rownames for our study data frame to be the patient identifiers. These are also used as basenames for the IDAT files storing the imprintome methylation data.
For example, the identifier "GSM8285531_207344530004_R01C01" refers to the red and green idat files for this patient.

1) GSM8285531_207344530004_R01C01_Grn.idat
2) GSM8285531_207344530004_R01C01_Red.idat

Please use this convention for you analysis since all of the downstream statisticall analysis assumes that rownames of study data will match column names in the output beta methylation matrix.

We can now load this data into tdhia and calculate the beta values and signal 
detection p-values. This is returned in a named list, with the following fields:  
**data$probe_beta_df**: dataframe of beta values, probe_id x sample_id.  
**data$probe_pval_df**: dataframe of signal p-values, probe_id x sample_id.  
**data$platform**: string that describes platform for methylation array.  
**data$manifest**: dataframe of the manifest file used for SeSame processing.  
  


```{r probes, eval=TRUE}


data_cache_path = file.path(local_download$directory, "data_cache")
dir.create(data_cache_path, showWarnings = F, recursive = T)

data <- tdhia::tdhia_pipeline(idat_dir_paths = file.path(local_download$directory, "idat"), 
                      OVERWRITE_TEMP_DATA = T, idat_basenames = study_metadata$Patient.ID,
                      probe_data_cache_path = file.path(data_cache_path, "AD_probe_data.rda"), 
                      multicore = T, discard_unmapped_probes = T, set_failed_betas_na = T, merge_replicates = "pre_beta",
                      
                      discard_failed_probes = T, db_flag = F)


```
  
You can access the probe beta matrix with the **probe_beta_df** field. The number of rows for this matrix will depend on whether merging replicates was specified. If `merge_replicates = NULL`, then the number of rows will match the number of rows in the array manifest. Otherwise probes are are replicates are merged together.

```{r visualize, eval=TRUE}
# Access probe beta matrix
head(data$probe_beta$probe_beta_df)

```


After importation, some probes are filtered out due to poor probe performance, patient data quality, or individual measurement quality thresholds. You can access the filtered probe data in *filt_probe_beta* field.
We can filter out low quality probes with the filt_probe_beta(). This function 
returns a named list with the same format as *probe_beta* above.  

```{r filter_visualize, eval=TRUE}
# Access filtered probe beta matrix
head(data$filt_probe_beta$probe_beta_df)
```



Once probe data has been filtered convert the probe beta matrix to a cpg data 
matrix, stored in a named list of the form:  
**data$cpg_beta_df**: dataframe of beta values, probe_id x sample_id.  
**data$platform**: string that describes platform for methylation array.  
**data$manifest**: dataframe of the manifest file used for SeSame processing.  
**data$cpg_n_probes**: number of probes for each cpg site.
  
Probe beta values that belong to the same cpg site for a given patient are averaged.  

You can access the filtered probe beta matrix with the **cpg_beta_df** field.
```{r cpg_visualize, eval=TRUE}
# Access filtered probe beta matrix
head(data$cpg_beta$cpg_beta_df)
```

# Generate QC Report

After the CpG matrix has been created, a QC report can be generated. This report
summarizes missingness and beta-value distributions by sample and feature, and
optionally saves a PDF with histograms, density overlays, a sample-sample
correlation heatmap, and PCA plots. When the relevant matrices are present, the
report also includes detection p-value QC, retention counts, design score
summaries, ICR coverage, replicate-probe agreement, and sample outlier flags.

**qc_cpg**: dataframe of beta values, probe_id x sample_id.  
**qc_cpg**: string that describes platform for methylation array.  
**qc_cpg**: dataframe of the manifest file used for SeSame processing.  
**qc_cpg**: number of probes for each cpg site.

```{r qc_report, eval=TRUE}
data_beta <- list(
  filt_probe_beta = data$filt_probe_beta,
  cpg_beta = data$cpg_beta
)

qc_cpg <- tdhia::tdhia_qc_report( data_beta = data_beta,
  output_dir_path = "tdhia_results", level = "cpg", show_plots = F,
  save_pdf = F, save_metrics = F)

qc_cpg$sample_metrics
qc_cpg$feature_metrics
```



Finally, we can convert the cpg beta matric to an icr beta matric, which is 
returned in a named list of the form:  
**icr_beta_df**: dataframe of beta values, icr_id x sample_id.  
**platform**: string that describes platform for methylation array.  
**manifest**: dataframe of the manifest file used for SeSame processing.  
**n_CpGs**: number of cpg sites in each icr site.
  
Probe beta values that belong to the same cpg site for a given patient are averaged.  

You can access the filtered probe beta matrix with the **data$probe_beta_df** field.
```{r eval=TRUE}
# Access filtered probe beta matrix
head(data$icr_beta$icr_beta_df)
```

# CpG Level Association Study {.tabset}

We can perform an association study between individual CpG sites in the imprintomeon Alzheimer disease status. We specify the response variable as disease state, the predictor as CpG methylation state, and adjust for Age, Sex, and ethnicity as covariates. This method of analysis is the weakest statically because it does not look for coordinated effects across ICRs, but there is more flexibility with this method for model design. Methylation state can be specified as a response variable or predictor, unlike the other approaches available for analysis. This approach also has support for imputation with MICE, and 

disease_state ~ cpg_methylation + age + sec + ethnicity

We analyze all patients initially, and then separate patients by ethnicity in case of nonlinear responses.


## All Patients


```{r association1, eval=TRUE,results = 'hide'}
set.seed(1)
n.cores <- max(c(parallel::detectCores()-2,1))

# disease_state ~ cpg + Age + Sex, All patients
df_all <- tdhia::imprintome_glm(
  model_str = "disease_state ~ beta + Age + Sex + Ethnicity", study_data = study_metadata,
  betas = data$cpg_beta$cpg_beta_df, family = "binomial", m_value_transform = T,
  n_p_adj = nrow(data$icr_beta$icr_beta_df), db_flag = FALSE, rm.na.all = FALSE, 
  verbose = TRUE, impute_na = TRUE, max_p_val = 0.05, n.cores = n.cores)
```

## Non-hispanic White
```{r association2, eval=TRUE,results = 'hide'}
# disease_state ~ cpg + Age + Sex, Non-hispanic white only
ids_white <- study_metadata$Patient.ID[study_metadata$Ethnicity == "NHW"]
df_white <- tdhia::imprintome_glm(
  model_str = "disease_state ~ beta + Age + Sex",
  study_data = study_metadata %>% filter(Patient.ID %in% ids_white),
  betas = data$cpg_beta$cpg_beta_df, family = "binomial", m_value_transform = T,
  n_p_adj = nrow(data$icr_beta$icr_beta_df), db_flag = FALSE, rm.na.all = FALSE, 
  verbose = TRUE, impute_na = TRUE, max_p_val = 0.05, n.cores = n.cores)
```


## Non-hispanic Black
```{r association3, eval=TRUE,results = 'hide'}
# disease_state ~ cpg + Age + Sex, Non-hispanic black patients
ids_black <- study_metadata$Patient.ID[study_metadata$Ethnicity == "NHB"]
df_black <- tdhia::imprintome_glm(
  model_str = "disease_state ~ beta + Age + Sex",
  study_data = study_metadata %>% filter(Patient.ID %in% ids_black),
  betas = data$cpg_beta$cpg_beta_df, family = "binomial", m_value_transform = T,
  n_p_adj = nrow(data$icr_beta$icr_beta_df), db_flag = FALSE, rm.na.all = FALSE, 
  verbose = TRUE, impute_na = TRUE, max_p_val = 0.05, n.cores = n.cores)


```


# ICR-level regression tests

TDHIA also includes regression tools for testing ICR-level methylation patterns.
Two direct ICR-level options are principal component regression and SKAT. A third
option first tests individual CpGs with LIMMA and then combines CpG-level
p-values within each ICR using Lancaster's method.

For binary traits, make sure the phenotype is coded as a two-level factor and
set the reference group before modeling. The examples below use the simulated
`smoked` variable as a placeholder binary outcome. For continuous outcomes, use
`family = "continuous"` in `pc_regression_test()` and `out_type = "C"` in
`skat_icr_test()`.

These tests do not currently support missing data, so we must reprocess the array data and not filter individual beta measurements with poor measurement quality `set_failed_betas_na = F` (although poor performing probes can still be filtered across all patients).
These tests also are restricted with the model design, where beta values can only be the response variable or predictor. Currently only the CpG-level analysis supports missing data imputation and beta values being a response or predictor.

```{r eval=TRUE, results = FALSE}
# Reprocess array data with 
data_no_na <- tdhia::tdhia_pipeline(idat_dir_paths = file.path(local_download$directory, "idat"), 
                      OVERWRITE_TEMP_DATA = F, idat_basenames = study_metadata$Patient.ID,
                      probe_data_cache_path = file.path(data_cache_path, "AD_probe_data_no_na.rda"), 
                      multicore = T,discard_unmapped_probes = T, set_failed_betas_na = F, 
                      discard_failed_probes = T, db_flag = F)

```

## Principal component regression

Principal component regression summarizes the CpGs inside each ICR with enough
principal components to explain the requested fraction of methylation variance.
It then compares a full model that includes those methylation PCs to a reduced
model with only the covariates.

This analysis method is not designed for small datasets, and leads to overfitting and overspecification. For this vignette, we limit the number of principle components to 1 (`n_pcs = 1`) and only use one of the covariates in order the for the code to run. For an actual analysis, we would use all covariates and set `n_pcs = 1`.

We recommend to set `m_value_transform = TRUE` in so beta values are transformed to M-values in order to stabilize the variance and make the data more homoscedastic.

```{r eval=TRUE}

pc_icr_results <- tdhia::pc_regression_test(
  cpg_beta = Matrix::t(data_no_na$cpg_beta$cpg_beta_df) %>% as.data.frame(),
  m_value_transform = TRUE,  data_norm_type = "n1", pct_variance = 0.80, n_pcs = 1,
  df_study = study_metadata %>% rename(Patient_ID = Patient.ID),
  outcome = "disease_state",  covariates = c("Sex"),
  Patient_ID = "Patient_ID",  family = "binomial",  icr_ids = NULL,
  min_cpg = 3,  verbose = TRUE,  n.cores = 1)

pc_icr_results <- pc_icr_results %>%
  dplyr::left_join(
    tdhia::add_metadata_to_imp_sites(pc_icr_results$ICR_id, imp_type = "icr", db_flag = FALSE),
    by = c("ICR_id" = "icr_id")
  )

head(pc_icr_results)
```

## SKAT ICR regression

SKAT tests each ICR as a set of CpG methylation values. It is useful when the
CpGs inside an ICR may have heterogeneous effects. A drawback to SKAT is there no effect size measure reported for ICRs, only statistical significance.

We recommend to set `m_value_transform = TRUE` in so beta values are transformed to M-values in order to stabilize the variance and make the data more homoscedastic.

```{r eval=TRUE}
skat_icr_results <- tdhia::skat_icr_test(
  cpg_betas = as.data.frame(data_no_na$cpg_beta$cpg_beta_df),
  df_study = study_metadata,  response = "disease_state",
  predictors = c("Sex", "Age", "Ethnicity"),  method = "optimal.adj",
  out_type = "D",  icr_ids = NULL,  min_cpg = 3,  db_flag = FALSE,
  m_value_transform = TRUE,  scaling = TRUE,  verbose = TRUE,
  n.cores = max(c(parallel::detectCores()-2, 1))
)

skat_icr_results <- skat_icr_results %>%
  dplyr::left_join(
    tdhia::add_metadata_to_imp_sites(skat_icr_results$icr_id, imp_type = "icr", db_flag = FALSE),
    by = "icr_id"
  )

head(skat_icr_results)
```

## LIMMA followed by Lancaster aggregation {.tabset}

Next we can perform a CpG-level DML analysis and ICR-level DMR analysis using a
LIMMA-based workflow.

The CPG level analyses uses LIMMA for increasing power for differential methylation 
testing. This example queries external annotation services through downstream
annotation helpers, so it is shown but not evaluated during package checks. Run
it interactively when network access and annotation resources are available.


```{r eval=TRUE}

cpg_dml_out <- tdhia::cpg_dml_test(
  df_study = study_metadata, predictors = c("disease_state", "Age", "Sex", "Ethnicity"), cpg_beta = data$cpg_beta$cpg_beta_df,
  pvalue_threshold = 0.001, db_flag = F, sample_name = "Patient.ID", correlation_check = F,
  beadchip_correction = F, verbose = T, write_plots = F,
  output_dir_path = getwd())


```

### QQ plot

```{r, include = TRUE, echo = TRUE}
cpg_dml_out$plots$qq_p_value
```

### Volcano plot

```{r, include = TRUE, echo = TRUE}
cpg_dml_out$plots$volcano
```

### Manhattan plot

```{r, include = TRUE, echo = TRUE}
cpg_dml_out$plots$manhattan
```

```{r, include = TRUE, echo = TRUE}
knitr::kable(cpg_dml_out$df_dml, format = "html") %>% kableExtra::kable_styling()
```


The ICR level analyses uses Lancaster p-value aggregation procedure to test for
altered methylation state in ICRs.


```{r eval=TRUE}

# Differential icr analysis
# ICR Lancaster test using LIMMA p-values
icr_dmr_test <-  tdhia::icr_dmr_test(df_dml = cpg_dml_out$df_dml, chr_lens = cpg_dml_out$chr_lens, 
                                     pval_threshold = 0.05,
                         fdr_sig_threshold = 0.0001,
                         verbose = T, db_flag = F)

icr_dmr_test$plots$manhplot_icr
```



## Package Information

This vignette was prepared for **tdhia** version 0.1.0. The package provides
helpers for importing TDHIA IDAT files, filtering probe-level methylation
signals, aggregating probes to CpG and ICR-level matrices, generating QC
summaries, and running CpG- and ICR-level methylation association workflows.

For reproducibility, record the package and R session versions used for each
analysis run.

```{r package_info, eval=TRUE}
packageVersion("tdhia")
sessionInfo()
```
